---
Title: Activity Data
Author: H.Smit
Date: 11/10/2021
Output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ACTIVITY DATA

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The variables included in this dataset are:

    steps: Number of steps taking in a 5-minute interval (missing values are coded NA)
    
    date: The date on which the measurement was taken in YYYY-MM-DD format
    
    interval: Identifier for the 5-minute interval in which measurement was taken.
    

### 1 ] READING & PREPROCESSING THE DATA

```{r activity}

# Load libraries
    library(readr)
    library(ggplot2)
    library(dplyr)

# Download file from web
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", destfile="activity.zip", 
                  mode="wb")

# Unzip data  
    unzip("activity.zip")
    activity <- read.csv("activity.csv", header = TRUE)
    activity$Date <- as.Date(activity$date)

```

###  2 ] HISTOGRAM OF TOTAL NUMBER OF STEPS TAKEN EACH DAY

```{r steps}
    
# Calculate the total number of steps for each day
    TotalStepsPerDay <- activity %>%
        group_by(Date) %>%
        summarize(sumsteps = sum(steps, na.rm = TRUE)) 

# Display first 10 rows of data
    head(TotalStepsPerDay,10)

# Histogram of Total steps per day
    hist(TotalStepsPerDay$sumsteps, main = "Histogram of Daily Steps", 
    col="purple", xlab="Steps", ylim = c(0,30))

```


###  3 ] MEAN AND MEDIAN NUMBER OF STEPS EACH DAY

```{r Mean Median}
# calculate the Mean and Median prior to removing NA
    MeanPreNA <- round(mean(TotalStepsPerDay$sumsteps),digits = 2)
    MedianPreNA <- round(median(TotalStepsPerDay$sumsteps),digits = 2)

# Print the result for Mean
    print(paste("The mean is: ", MeanPreNA))

# Print the result for Median
    print(paste("The median is: ", MedianPreNA))
```



### 4 ] TIME SERIES PLOT OF THE AVERAGE NUMBER OF STEPS TAKEN

```{r Time Series}

# Calculate the Steps per Interval
     TotalStepsPerInterval <- activity %>%
        group_by(interval) %>%
        summarize(meansteps = mean(steps, na.rm = TRUE)) 

# Display first 10 rows of data
     head(TotalStepsPerInterval,10)

# Plot Steps per Interval
     plot(TotalStepsPerInterval$meansteps ~ TotalStepsPerInterval$interval,
     col="blue", type="l", xlab = "5 Minute Intervals", ylab = "Average Number of Steps",
     main = "Steps By Time Interval")
```



### 5 ] 5-MINUTE INTERVAL CONTAINS MAXIMUM NUMBER OF STEPS

```{r Max Interval}
# Print the maximum number of steps
    print(paste("Interval containing the most steps on average:   ",TotalStepsPerInterval$interval[which.max(TotalStepsPerInterval$meansteps)]))

# Print the average number of steps per interval
    print(paste("Average steps for that interval: ",round(max(TotalStepsPerInterval$meansteps),digits=2)))


```


### 6 ] CODE TO DESCRIBE AND SHOW STRATEGY FOR IMPUTTING MISSING DATA

```{r Missing data}

# Calculate the amount of missing data
    print(paste("The total number of rows with NA is: ",sum(is.na(activity$steps))))

# Strategy for filling in all of the missing values in the dataset
# To solve for the missing NA values the average for the associated interval will be used. The average was built 
# in a prior step and is readily available: 



# Create a new dataset that is equal to the original dataset but with the missing data filled in

# Dataset before imputing missing values

# Display first 10 rows of data
    head(activity,10)

# First, loop through all records of a copy of the ‘activity’ data. Then, look for records containing NA values. Finally, transform the ‘steps’ value based on matching the interval in the ‘TotalStepsPerInterval’ data frame created in a prior step
    
    activityNoNA <- activity  
        for (i in 1:nrow(activity)){
            if(is.na(activity$steps[i])){
                activityNoNA$steps[i]<- TotalStepsPerInterval$meansteps[activityNoNA$interval[i] ==
                                                                          TotalStepsPerInterval$interval]
              }
          }

# Dataset after imputing missing values
    
# Display first 10 rows of data
    head(activityNoNA,10)
```


### 7 ] HISTOGRAM OF TOTAL NUMBER OF STEPS TAKEN EACH DAY AFTER MISSING VALUES ARE IMPUTED

```{r No missing Values}
# Calculate the steps per day with imputing missing data
    TotalStepsPerDay <- activityNoNA %>%
        group_by(Date) %>%
        summarize(sumsteps = sum(steps, na.rm = TRUE)) 

# Show first 10 entries of table
    head(TotalStepsPerDay,10)

# Show histogram of data
    hist(TotalStepsPerDay$sumsteps, main = "Histogram of Daily Steps", 
     col="purple", xlab="Steps")

# calculate the mean after imputing data
    meanPostNA <- round(mean(TotalStepsPerDay$sumsteps), digits = 2)
    
# calculate the median after imputing data
    medianPostNA <- round(median(TotalStepsPerDay$sumsteps), digits = 2)

# Print the mean and median after imputing data
    print(paste("The mean is: ", mean(meanPostNA)))

    print(paste("The median is: ", median(medianPostNA)))

# Create table compare data before and after imputing data
    NACompare <- data.frame(mean = c(MeanPreNA,meanPostNA),median = c(MedianPreNA,medianPostNA))
    rownames(NACompare) <- c("Pre NA Transformation", "Post NA Transformation")
    print(NACompare)

```
### 8 ] PANEL PLOT COMPARING THE AVERAGE NUMBER OF STEPS TAKEN PER 5-MINUTE INTERVAL ACROSS WEEKDAYS AND WEEKENDS

```{r Compare}

# Create a new factor variable in the dataset with two levels - “weekday” and “weekend” indicating whether a given date  is a  weekday or weekend day.

    activityDoW <- activityNoNA
    activityDoW$date <- as.Date(activityDoW$date)
    activityDoW$day <- ifelse(weekdays(activityDoW$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
    activityDoW$day <- as.factor(activityDoW$day)

# Make a panel plot containing a time series plot (i.e. type=“l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
    activityWeekday <- filter(activityDoW, activityDoW$day == "weekday")
    activityWeekend <- filter(activityDoW, activityDoW$day == "weekend")

# Summary of Weekday    
    activityWeekday <- activityWeekday %>%
        group_by(interval) %>%
        summarize(steps = mean(steps)) 
    activityWeekday$day <- "weekday"
    
# Summary of weekend
    activityWeekend <- activityWeekend %>%
        group_by(interval) %>%
        summarize(steps = mean(steps)) 
    activityWeekend$day <- "weekend"

    wkdayWkend <- rbind(activityWeekday, activityWeekend)
    wkdayWkend$day <- as.factor(wkdayWkend$day)

# Create a ggplot
    g <- ggplot (wkdayWkend, aes (interval, steps))
    g + geom_line(col = "blue") + facet_grid (day~.) + 
        theme(axis.text = element_text(size = 12),axis.title = element_text(size = 14)) + 
        labs(y = "Number of Steps") + labs(x = "Interval") + 
        ggtitle("Average Number of Steps - Weekday vs. Weekend") + 
        theme(plot.title = element_text(hjust = 0.5))


```

There is differences between activity when comparing weekday vs weekends.  People might exercise more during weekend versus weekday during the day, while in the week, people are most likely to exercise in early in the morning vs late afternoon.
